The role of cognition and movement of seed-dispersing animals in shaping plant distributions

### 2023-08-01 Text/Article_ModelSeedDispersal.Rmd
Update for submission

---
title: "The role of seed-dispersing animals in shaping plant distributions"
# "Animal cognition as a cause and consequence of resource distribution heterogeneity"
# "Eco-evolutionary feedback loops between animal cognition, movement and resource distribution"
# The role of animal cognition in shaping resource distribution via zoochory
# date: "`r format(Sys.time(), '%B %d, %Y')`"

@@

\faTags \hspace{0.01in} **Keywords:** Agent-based model - Feedback loop - Foraging - Frugivory - Memory - Movement - Phenology - Seed dispersal

@@

\initial{\textcolor{black}{H}}ypotheses about the evolution of vagile plant-eating foragers' spatiotemporal cognition (i.e. the ability to process, integrate and utilise information on resource distribution and phenology) generally assume that foragers evolve in an unchanging environment [@rosati2017foraging]. However, while foragers' cognition is likely to be shaped by the distribution of resources [@milton1981distribution; @grove2013evolution; @robiratemporalmemory; @bracis2015memory; @boyer2010modelling], the characteristics of the resources themselves could be the result of forager-induced selection [@lomascolo2010signal; @rojas2021combination]. In the tropical rainforests, for example, the cognition of frugivorous primates has probably been shaped by the challenge of the ephemerality and dispersion of their food [@milton1981distribution; @zuberbuhler2010foraging; @trapanese2019review], supporting complex [@teichroeb2019navigation; @trapanese2019whatwhere; @janson2016capuchins; @janson1998experimental; @janmaat2013mistake; @robirarandomwalk] and flexible [@trapanese2022flexibility; @janmaat2006primates] movement heuristics. In turn, primates remember and forage for the most rewarding trees [@ban2016fattyfruit; @ban2014highreward; @florchinger2010fruit]. The latter are thus more likely to have their seeds dispersed, driving plant and landscape evolution [@chapman2013primates; @lambert1998evolutionary].

@@

### Agent

@@

$I = \frac{1}{\sum_{ij}{w_{ij}}}\frac{\sum_{ij}w_{ij}d(x_i,\bar{t})d(x_j,\bar{t})}{N^{-1}\sum_i d(x_i,\bar{t})^2}$

@@

```{r figEngineering, fig.width=19.25, fig.height=19.25, include=TRUE, warning = FALSE, fig.cap="Seed-dispersing foragers can shape resource distribution | The plots show the results of 200 simulations when a seed-dispersing forager with varying level of cognition (i.e. proportion of tree locations and associated phenology known), moving choice rules or space competition intensities between trees (the two latter being simulated with an omniscient forager; left to right) was let foraging for a long time. The resulting distribution pattern was assessed by its patchiness (the tendency of trees to be aggregated into patches; larger values indicate highly aggregated trees), its alignement (the tendency of trees to form linear geometries; larger values indicate the presence of linear geometries) and spatial autocorelation in fruiting date (the tendency of trees to have fruiting dates closer/farther in time to their spatial neighbours; values close to 0 indicate negative spatial autocorrelation, values close to 1 indicate no spatial autocorrelation, values above 1 indicate positive spatial autocorrelation). The emerging movement behaviour of the forager was described in terms of routine behaviour (the tendency of the forager to target trees in a predictible way; values close to 0 indicate no predictible movement, values close to 1 indicate highly predictible movements)."}

@@

plotEfficiency +
  geom_segment(meanValuesEfficiency,
               mapping = aes(x = initCondition, xend = initCondition, y = minY + 0.075, yend = maxY - 0.075),
               position = positionChange,
               arrow = arrow(length = unit(0.02, "npc"), ends="last"),
               color = "gray",
               size = 0.4) +
  geom_text(meanValuesEfficiency,
               mapping = aes(x = initCondition,  y = minY + (maxY - minY)/2, label = text),
               position = positionChange,
               color = "gray",
               hjust = -0.1,
               size = 3) +
  scale_fill_manual(name = "forager type", values = c("white", "grey80", "black")) #Change the colours

@@

|    As the breadth of knowledge increased, routine behaviour, the ability to repeatedly follow a fixed sequence of movements, developed [routine index > 0.8, @riotte2015memory]. The model evidenced that the emergence of such behaviour was associated with changes in the distribution of resources, which tended to become more patchy, eventually forming linear features paralleling the routes used. The agglomeration of seeds for cognitive foragers has been proposed theoretically [@john2016memoryseed] and empirically [@fragoso2003tapiraggregationseed], ultimately impacting the genetic structure and flow of plant populations [@gelmicandusso2017zoochorygenetic]. Here, the model highlighted how engineering can go further by contributing to a more organised distribution of resources, with the emergence of linear features likely to be associated with the foraging routes used due to cognition, as suggested by forest elephant routes matching with fruit gradients in the forest [@blake2009forest]. Indeed, the non-linearity of the changes in patchiness and alignment with the cognition of the forager likely corresponds to the formation of the routes themselves, with the "materialisation" first of the route intersections (hence higher patchiness, lower alignment for low cognition), where most of the recursions occur [@presotto2018intersection], before a spreading along the segment path connecting them is possible when cognition increases. This may explain why animal-dispersed plant species are less aggregated than when other dispersal mechanisms are involved, such as gravichory, where dispersed seeds do not travel far from the parent tree [@seidler2006seed]. This engineering remained yet limited because the resource patterns did not fully match the simulated route-based patterns [(see Appendix)][Using patchiness and alignment to characterise spatial point patterns]. Furthermore, despite vertical inheritance of the parent tree phenology, seed dispersal was not sufficient to drive local autocorrelation in fruiting dates. Tree synchrony is an important tool that can be used by foragers to plan their movements efficiently [@janmaat2012use]. Given the climatic drivers of large-scale synchrony, the latter can be tracked by animals [@janmaat2006primates]. A fine-scale synchrony is also observed in nature [@lamontagne2007local] and might as likely affect animal foraging success. While large-scale population trends in phenology might stem from genetic constraints [@chuine2000geneticphenology; @lyngdoh2012influence], the absence of emergence of spatial autocorrelation in fruiting date in this model may suggest a mainly non-genetic origin of these fine-scale variations for endozoochory (as implicitly assumed here) given the delay between ingestion and defecation, in the absence of other dispersal modes [e.g. gravichory, leading to highly clustered patterns, @seidler2006seed]. However, these conclusions may be challenged by more accurate modelling of movement behaviour, which may lead to different results (see below). Despite these limits, the observed spatial engineering could still be sufficient for the emergence of eco-evolutionary feedback loops between animal resources and animal movement (and cognition) [@riotte2020environmental] due to the evidenced repercussions on foragers' foraging efficiency.

@@

|   Space engineering was reinforced by the conjunction of two forces: the forager movement rules and competition for space between trees. Maximum resource engineering was achieved when the individual moved to all the trees it encountered along the way. At the other extreme, when the animal moved directly to the best target, engineering was reduced, following routine trends. Although more efficient, moving straight to the target is rarely observed in nature. Instead, animals tend to move along routes, sometimes bypassing or otherwise opportunistically inspecting trees without resources, fitting with a beacon-based navigation, where individuals move from beacon to beacon to reach a goal [@warren2019non; @fagan2013spatial]. Such a movement strategy gives them the possibility to monitor the environment [@guinea2019arboreal], allowing them to update personal information. As such, despite short-term costs due to detours, moving to all tress might provide long-term advantages. Simulated movement patterns remain extremely simplistic and animal movement is much more nuanced than in this modelling work. This could be important for seed acquisition and deposition [@morales2022models]. This model considered directed movement supported by memory, which could alternate with random steps when memory was limited but independent of local cue feedback, as it would rather be expected for efficient foraging [@benhamou2015ultimate]. In reality, animal movement may indeed alternate between 'large-scale' directed relocation events and 'area-restricted' wandering searches of the tree in a non-random manner [@bartumeus2016foraging]. When and where area-restricted searches occur may further contribute to the emergent distribution of seed deposition [@westcott2005incorporating; @russo2006behaviourdispersal], provided that both seed deposition and foraging can occur simultaneously. Animals may yet be selective about where and when they defecate. The presence of latrines may contribute to shaping the distribution of resources too [@irwin2004lemurlatrine; @fragoso1997tapir; @pouvelle2009howlerlatrine; @fuzessy2022defecation; @sakamoto2015raccoon]. Overall, as movement and/or dispersal scales with behaviour/personality [@zwolak2018intraspecificdispersal; @brehm2022personalitydispersal; @zwolak2020personalitydispersal], but also species and individual size [@jordano2007frugivores], a multi-individual and multi-species system could lead to a more complex multi-scale scaffolding than illustrated here. This could explain why the evolution of the cognitive machinery underpinning foraging might depend on the frugivore community composition [@robira2023primate].

@@

```{r figShrinkage, fig.width=20, fig.height=5, include=TRUE, warning = FALSE, fig.cap="Shrinkage of resource range is driven by cognition, the forager's movement rules and competition for space between trees."}
shrinkageMovingRule_plot <- readRDS("../Renvironment/Plots/movingRuleShrinkage.rds")
shrinkageSpaceTree_plot <- readRDS("../Renvironment/Plots/spacingTreeShrinkage.rds")
shrinkageMain_plot <- readRDS("../Renvironment/Plots/mainShrinkage.rds")

@@

>>>>

---
title: "The role of cognition and movement of seed-dispersing animals in shaping plant distributions"
# "Animal cognition as a cause and consequence of resource distribution heterogeneity"
# "Eco-evolutionary feedback loops between animal cognition, movement and resource distribution"
# The role of animal cognition in shaping resource distribution via zoochory
# date: "`r format(Sys.time(), '%B %d, %Y')`"

@@

\faTags \hspace{0.01in} **Keywords:** Agent-based model - Feedback loop - Foraging - Frugivory - Memory - Movement - Phenology - Zoochory

@@

\initial{\textcolor{black}{H}}ypotheses about the evolution of vagile plant-eating foragers' spatiotemporal cognition (i.e. the ability to process, integrate and utilise information on resource distribution and phenology) generally assume that foragers evolve in an unchanging environment [@rosati2017foraging]. However, while foragers' cognition is likely to be shaped by the distribution of resources [@milton1981distribution; @grove2013evolution; @robiratemporalmemory; @bracis2015memory; @boyer2010modelling], the characteristics of the resources themselves could be the result of forager-induced selection [@lomascolo2010signal; @rojas2021combination]. In the tropical rainforests, for example, the cognition of frugivorous primates has probably been shaped by the challenge of the ephemerality and dispersion of their food [@milton1981distribution; @zuberbuhler2010foraging; @trapanese2019review], supporting complex [@teichroeb2019navigation; @trapanese2019whatwhere; @janson2016capuchins; @janson1998experimental; @janmaat2013mistake; @robirarandomwalk] and versatile [@trapanese2022flexibility; @janmaat2006primates; @robira2023seasonal] movement heuristics. In turn, primates remember and forage for the most rewarding trees [@ban2016fattyfruit; @ban2014highreward; @florchinger2010fruit]. The latter are thus more likely to have their seeds dispersed, driving plant and landscape evolution [@chapman2013primates; @lambert1998evolutionary].

@@

### Forager

@@

$I = \frac{1}{\sum_{ij}{w_{ij}}}\frac{\sum_{ij}w_{ij}d(t_i,\bar{t})d(t_j,\bar{t})}{N^{-1}\sum_i d(t_i,\bar{t})^2}$

@@

```{r figEngineering, fig.width=19.5, fig.height=19.5, include=TRUE, warning = FALSE, fig.cap="Seed-dispersing foragers can shape resource distribution | The plots show the results of 200 simulations when a seed-dispersing forager with varying level of cognition (i.e. proportion of tree locations and associated phenology known), moving choice rules or space competition intensities between trees (the two latter being simulated with an omniscient forager; left to right) was let foraging for a long time. The resulting distribution pattern was assessed by its patchiness (the tendency of trees to be aggregated into patches; larger values indicate highly aggregated trees), its alignement (the tendency of trees to form linear geometries; larger values indicate the presence of linear geometries) and spatial autocorelation in fruiting date (the tendency of trees to have fruiting dates closer/farther in time to their spatial neighbours; values close to 0 indicate negative spatial autocorrelation, values close to 1 indicate no spatial autocorrelation, values above 1 indicate positive spatial autocorrelation). The emerging movement behaviour of the forager was described in terms of routine behaviour (the tendency of the forager to target trees in a predictible way; values close to 0 indicate no predictible movement, values close to 1 indicate highly predictible movements)."}

@@

plotEfficiency +
  geom_segment(meanValuesEfficiency,
               mapping = aes(x = initCondition, xend = initCondition, y = minY + 0.075, yend = maxY - 0.075),
               position = positionChange,
               arrow = arrow(length = unit(0.02, "npc"), ends="last"),
               color = "gray",
               size = 0.4) +
  geom_text(meanValuesEfficiency,
               mapping = aes(x = initCondition,  y = minY + (maxY - minY)/2, label = text),
               position = positionChange,
               color = "gray",
               hjust = -0.1,
               size = 3) +
  scale_fill_manual(name = "Forager type", values = c("white", "grey80", "black")) #Change the colours

@@

|    As the breadth of knowledge increased, routine behaviour, the ability to repeatedly follow a fixed sequence of movements, developed [routine index > 0.8, @riotte2015memory]. The model evidenced that the emergence of such behaviour was associated with changes in the distribution of resources, which tended to become more patchy, eventually forming linear features paralleling the routes used. The agglomeration of seeds for cognitive foragers has been proposed theoretically [@john2016memoryseed] and empirically [@fragoso2003tapiraggregationseed], ultimately impacting the genetic structure and flow of plant populations [@gelmicandusso2017zoochorygenetic]. Here, the model highlighted how engineering can go further by contributing to a more organised distribution of resources, with the emergence of linear features likely to be associated with the foraging routes used due to cognition, as suggested by forest elephant routes matching with fruit gradients in the forest [@blake2009forest]. Indeed, the non-linearity of the changes in patchiness and alignment with the cognition of the forager likely corresponds to the formation of the routes themselves, with the "materialisation" first of the route intersections (hence higher patchiness, lower alignment for low cognition), where most of the recursions occur [@presotto2018intersection], before more routes are used and formed when cognition increases. Long-distance travels through these routes may explain why animal-dispersed plant species are less aggregated than when other dispersal mechanisms are involved, such as gravichory, where dispersed seeds do not travel far from the parent tree [@seidler2006seed]. This engineering remained yet limited because the resource patterns did not fully match the simulated route-based patterns [(see Appendix)][Using patchiness and alignment to characterise spatial point patterns]. Furthermore, despite vertical inheritance of the parent tree phenology, seed dispersal was not sufficient to drive spatial autocorrelation in fruiting dates. This was not because the trees all ended up fruiting at the same time, as there was still ample variation in fruiting timing at the end of the simulations [(see Appendix)][Variations in fruiting dates]. Tree synchrony is an important tool that can be used by foragers to plan their movements efficiently [@janmaat2012use]. Given the climatic drivers of large-scale synchrony, the latter can be tracked by animals [@janmaat2006primates]. A fine-scale synchrony is also observed in nature [@lamontagne2007local] and might as likely affect animal foraging success. While large-scale population trends in phenology might stem from genetic constraints [@chuine2000geneticphenology; @lyngdoh2012influence], the absence of emergence of spatial autocorrelation in fruiting date in this model (which considers vertical inheritance) may suggest a mainly non-genetic origin of these fine-scale variations in the absence of other dispersal modes [e.g. gravichory, leading to highly clustered patterns, @seidler2006seed]. However, these conclusions may be challenged by more accurate modelling of movement behaviour, which may lead to different results (see below). Despite these limits, the observed spatial engineering could still be sufficient for the emergence of eco-evolutionary feedback loops between animal resources and animal movement (and cognition) [@riotte2020environmental] due to the evidenced repercussions on foragers' foraging efficiency.

@@

|   Space engineering was reinforced by the conjunction of two forces: the forager movement rules and competition for space between trees. Maximum resource engineering was achieved when the individual moved to all the trees it encountered along the way. At the other extreme, when the animal moved directly to the best target, engineering was reduced, following routine trends. Although more efficient, moving straight to the target is rarely observed in nature. Instead, animals tend to move along routes, sometimes bypassing or otherwise opportunistically inspecting trees without resources, fitting with a beacon-based navigation, where individuals move from beacon to beacon to reach a goal [@warren2019non; @fagan2013spatial]. Such a movement strategy gives them the possibility to monitor the environment [@guinea2019arboreal], allowing them to update personal information. As such, despite short-term costs due to detours, moving to all tress might provide long-term advantages. However, the simulated movement patterns remain extremely simplistic. Animal movement is in reality much more nuanced than in this modelling work. This could be important for seed acquisition and deposition [@morales2022models]. For example, this model considered directed movement supported by memory, which could alternate with random steps when memory was limited but independent of local cue feedback, as it would rather be expected for efficient foraging [@benhamou2015ultimate]. Animal movement may indeed alternate between 'large-scale' directed relocation events and 'area-restricted' wandering searches of the tree in a non-random manner [@bartumeus2016foraging]. When and where area-restricted searches occur may further contribute to the emergent distribution of seed deposition [@westcott2005incorporating; @russo2006behaviourdispersal], provided that both seed deposition and foraging can occur simultaneously. Animals may yet be selective about where and when they defecate. The presence of latrines may contribute to shaping the distribution of resources too [@irwin2004lemurlatrine; @fragoso1997tapir; @pouvelle2009howlerlatrine; @fuzessy2022defecation; @sakamoto2015raccoon]. Overall, as movement and/or dispersal scales with behaviour/personality [@zwolak2018intraspecificdispersal; @brehm2022personalitydispersal; @zwolak2020personalitydispersal], but also species and individual size [@jordano2007frugivores], a multi-individual and multi-species system could lead to a more complex multi-scale scaffolding than illustrated here. This could explain why the evolution of the cognitive machinery underpinning foraging might depend on the frugivore community composition [@robira2023primate].

@@

```{r figShrinkage, fig.width=20, fig.height=8, include=TRUE, warning = FALSE, fig.cap="Shrinkage of resource range is driven by cognition, the forager's movement rules and competition for space between trees."}
shrinkageMovingRule_plot <- readRDS("../Renvironment/Plots/movingRuleShrinkage.rds")
shrinkageSpaceTree_plot <- readRDS("../Renvironment/Plots/spacingTreeShrinkage.rds")
shrinkageMain_plot <- readRDS("../Renvironment/Plots/mainShrinkage.rds")

@@

## Variations in fruiting dates

I quantified the variation in fruiting dates as one minus the length of the mean vector of fruiting dates [@batschelet1981circular] in either the initial or final conditions. It varies from 0 (no variation; all fruiting dates are the same) to 1 (extensive variation; fruiting dates are uniformly distributed). At the beginning or end of the simulations, the variation in fruiting dates was considerable, regardless of the scenario tested (Figure \@ref(fig:figVariationFruitingDates)).

```{r figVariationFruitingDates, fig.width=20, fig.height=8, include=TRUE, warning = FALSE, fig.cap=" Variation in fruiting dates at the beginning of the simulation ('initial' condition) or at the end ('final' condition) | The plots show the results of 200 simulations when a seed-dispersing forager with varying level of cognition (i.e. proportion of tree locations and associated phenology known), moving choice rules or space competition intensities between trees (the two latter being simulated with an omniscient forager; left to right) was let foraging for a long time. The index  of variation in fruiting dates ranges from 0 (no variation) to 1 (extensive variation)."}

plotVariationFruitingDates <- readRDS("../Renvironment/Plots/plotVariationFruitingDates.rds")
plotVariationFruitingDates

# References

### 2023-08-01 Text/coverLetter.Rmd
Update for submission

```{r}
journal <- "Peer Community In Ecology"
editor <- "Recommenders of PCI"
title <- "The role of seed-dispersing animals in shaping plant distributions"
listOfAuthors <- paste("Benjamin Robira", collapse = ", ")
```

>>>>

```{r}
journal <- "Peer Community In Ecology"
editor <- "Recommenders of PCI"
title <- "The role of cognition and movement of seed-dispersing animals in shaping plant distributions"
listOfAuthors <- paste("Benjamin Robira", collapse = ", ")
```
